### 2024-10-13 Article/main.tex
lemmi vari

\begin{prop}\label{prop:ext_left_to_funct}
    In \Cref{def:right_adjoint}, the map that takes an object $A$ to an object $L(A)$ can be extended to a functor $L: \cat{D \rightarrow C}$. Moreover, there exists a natural transformation $id_{\cat{D}}\to R\circ L$.
\end{prop}
\begin{proof}
    Let $R$ be the right adjoint as in ~\Cref{def:right_adjoint}. Given $f\colon X \rightarrow Y$, we can define $L(f)$ as the unqiue arrow  $L(X) \rightarrow L(Y)$ whose image through $R$ fits in the diagram below.
    \[
    \begin{tikzcd}
    	X \arrow[r, "{\eta_X}"] \arrow[d, "f" swap] & R(L(X)) \arrow[d, dashed, "R(L(f))"] \\
    	Y\arrow[r, "\eta_Y" swap]& R(L(Y))
    \end{tikzcd}
    \]

@@

In the category of graphs with equivalences, universal constructions are easy to obtain.
\begin{prop}COME SI CALCOLANO I LIMITI IN $\mathbf{EqGrph}$
\end{prop}
\begin{proof}
	TO DO
\end{proof}

 \todo{riscrivere questa parte come esempi che derivano dalla proposizione precedente}
The terminal object in $\mathbf{EqGrph}$ is the graph with only one node, only one edge and with the single node equivalent only to itself, while the initial object is the graph with no nodes and no edges.
Given two graphs with equivalences $\eqgraph{G} = (G, \sim_G) $ and $\eqgraph{H} = (H, \sim_{H})$, one can construct the product graph $\eqgraph{G \times H} = (G \times H, \sim_{G \times H})$, together with the two projections $\pi_G$ and $\pi_H$,  where $\sim_{G \times H}$ is the least equivalence relation on $G \times H$ such that, if $w_1$ and $w_2$ are nodes of $\eqgraph{G \times H}$, we have $w_1 \sim_{G \times H} w_2$ if $\pi_G(w_1) \sim_G \pi_G(w_2)$ and $\pi_H(w_1) \sim_H \pi_H(w_2)$.
An equalizer in $\mathbf{EqGrph}$ for a pair of morphisms $f, g: \eqgraph{G \rightarrow H}$ is exactly the same as in $\mathbf{Graph}$, namely, $(\eqgraph{E} = (E, \sim_E), e)$, where $E$ is the greatest subgraph of the underlying graph of $\eqgraph{G}$ on which $f$ and $g$ are equal, together with the inclusion $e: \eqgraph{E \rightarrow G}$. The equivalence relation is preserved ($\sim_E$ is $\sim_G$ restricted to the set of vertices of $E$).

@@

\section{Graphs as Functors}
\todo{non ha senso avere questa parte come sezione: "graph as functors"  ha
 senso come remark separato e la restante parte della sezione si salda a quella precedente}
Set-theoretical definitions of graphs are straightforward, but a further level of abstraction allows us to highlight some interesting properties of these objects. We can indeed consider a graph as a functor from a category of ``primitive objects'' (e.g., the object representing edges and the object representing vertices) onto a category which provides a representation of them. This concept is perfectly represented by categories of presheaves, where an object is nothing more than a way to interpret objects of a category as sets. An example will make clear what we are talking about.

@@

    A right adjoint of $Q$ is the functor $T: \mathbf{Graph \rightarrow EqGrph}$ \todo{stesso problema di prima. Io riscriverei anche questa metà perché non si capisce la notazione e ad un certo punto occorre usare le proprietà dei quozienti in Set, cosa che non è evidente.}such that, for each graph $\mathcal{G}$ of $\mathbf{Graph}$, $T(\mathcal{G}) = (\mathcal{G}, \approx_{\mathcal{G}})$, where $\approx_{\mathcal{G}}$ is the total relation among vertices of $\mathcal{G}$. We have then the following situation:
    \[
        \begin{tikzcd}
            \eqgraph{G} = (\mathcal{G}, \sim_{\mathcal{G}}) \arrow[r, "{\eta_{\eqgraph{G}}}"] \arrow[dr, "f"swap] & (T \circ Q)(\eqgraph{G}) = (\mathcal{G}/_{\sim_{\mathcal{G}}}, \approx_{\mathcal{G}/_{\sim_{\mathcal{G}}}}) \arrow[d, dashed, "h"] \\
            & T(\mathcal{H}) = \eqgraph{H}
        \end{tikzcd}
    \]
    Since $\eqgraph{H}$ is in the image of $T$, it is a graph with all equivalent nodes. Then, the morphism $f$ is a graph morphism that in addition sends each equivalence class of vertices onto the unique class in the graph $\eqgraph{H}$. Then, $f$ factors uniquely as $h \circ \eta_\eqgraph{G}$, and $h$ is a graph morphism extended to a morphism in $\mathbf{EqGrph}$ by sending the unique equivalence class on $(T \circ G)(\eqgraph{G})$ onto the unique equivalence class in $\eqgraph{H}$, and $\eta: Id_{\mathbf{EqGrph}} \rightarrow (T \circ Q)$ is the unit of the adjunction.
\end{proof}

@@

\begin{proof}
    Consider the following pullback in $\mathbf{EqGrph}$, where $m$ is a regular monomorphism:
    \[
        \begin{tikzcd}
            G \arrow[r, "m"] \arrow[d, "n" swap] & G_1 \arrow[d, "i"] \\
            G_2 \arrow[r, "j" swap] & P
        \end{tikzcd}
    \]
    In particular, since $\mathbf{EqGrph}$ is a category of presheaves, colimits are computed pointwise,\color{green}{commentato}\color{black}

    %%% IDEA: far vedere che, poiché i colimiti in una categoria di prefasci sono calcolati punto punto, allora ho un pushout su ogni componente (E, V, Q). Dalla prop nella parte sull'adesività si può concludere che tutti e tre i pushout preservano mono e quindi mono è preservata su tutte e tre le componenti, quindi reg  mono.
\end{proof}

}\color{black}

COSE DA FARE

>Quoziente crea i limiti (questo dovrebbe derivare direttamente dalla sezione precedente )

>Quoziente crea certi pushout (questo va fatto ex novo)

Regular monos in $\mathbf{EqGrph}$ are then natural transformations $\eta: \eqgraph{G \dot\rightarrow H}$ such that $\eta_Q$ is mono (\Cref{prop:reg_mono_in_EG_are_mono_in_graph}). Pullbacks preserve regular monomorphisms (\Cref{prop:monos_pres_by_pullback}). Regular monos are preserved also by pushouts (\Cref{lemma:reg_mono_pres_by_pushouts}). Again, since (co)limits are computed pointwise, and in $\Set$ monos are preserved by pushouts (since $\Set$ is adhesive), we can conclude what follows.

@@

@@

COSE DA FARE:
Questa sezione ha più o meno gli stessi problemi della precedente. L'ordine da rispettare imho è il seguente:

>Definizione
>Calcolo dei limiti e certi colimiti (si fanno come in EqGrph)
> cor del calcolo dei limiti: caratterizzare mono regolari
>I crea limiti e i giusti pushout
> e-graph sono quasiadesivi

>>>>

\begin{prop}\label{prop:ext_left_to_funct}
    In \Cref{def:right_adjoint}, the map that takes an object $A$ to an object $L(A)$ can be extended to a functor $L: \cat{D \rightarrow C}$. Moreover, there exists a natural transformation $id_{\cat{D}}\to R\circ L$.
\end{prop}
\todo{Devi mostrare che eta risulta naturale (viene in automatico ma conviene scriverlo per la dimostrazione successiva).}
\begin{proof}
    % Let $R$ be the right adjoint as in ~\Cref{def:right_adjoint}. It is possible to define, for $f: X \rightarrow Y$, $L(f) : L(X) \rightarrow L(Y)$  such that: \[R(L(f)) \circ \eta_X = \eta_Y \circ f\] \todo{finire}
    Let $R$ be the right adjoint as in ~\Cref{def:right_adjoint}. Given $f\colon X \rightarrow Y$, we can define $L(f)$ as the unqiue arrow  $L(X) \rightarrow L(Y)$ whose image through $R$ fits in the diagram below.
    \[
    \begin{tikzcd}
    	X \arrow[r, "{\eta_X}"] \arrow[d, "f" swap] & R(L(X)) \arrow[d, dashed, "R(L(f))"] \\
    	Y\arrow[r, "\eta_Y" swap]& R(L(Y))
    \end{tikzcd}
    \]

@@

In the category of graphs with equivalences, universal constructions are easy to obtain. 
\begin{prop}
    COME SI CALCOLANO I LIMITI IN $\mathbf{EqGrph}$
\end{prop}
\begin{proof}
    TO DO
\end{proof}
 \todo{riscrivere questa parte come esempi che derivano dalla proposizione precedente}
The terminal object in $\mathbf{EqGrph}$ is the graph with only one node, only one edge and with the single node equivalent only to itself, while the initial object is the graph with no nodes and no edges.
Given two graphs with equivalences $\eqgraph{G} = (G, \sim_G) $ and $\eqgraph{H} = (H, \sim_{H})$, one can construct the product graph $\eqgraph{G \times H} = (G \times H, \sim_{G \times H})$, together with the two projections $\pi_G$ and $\pi_H$,  where $\sim_{G \times H}$ is the least equivalence relation on $G \times H$ such that, if $w_1$ and $w_2$ are nodes of $\eqgraph{G \times H}$, we have $w_1 \sim_{G \times H} w_2$ if $\pi_G(w_1) \sim_G \pi_G(w_2)$ and $\pi_H(w_1) \sim_H \pi_H(w_2)$.
An equalizer in $\mathbf{EqGrph}$ for a pair of morphisms $f, g: \eqgraph{G \rightarrow H}$ is exactly the same as in $\mathbf{Graph}$, namely, $(\eqgraph{E} = (E, \sim_E), e)$, where $E$ is the greatest subgraph of the underlying graph of $\eqgraph{G}$ on which $f$ and $g$ are equal, together with the inclusion $e: \eqgraph{E \rightarrow G}$. The equivalence relation is preserved ($\sim_E$ is $\sim_G$ restricted to the set of vertices of $E$).

@@

\section{Graphs as Functors}

\todo{non ha senso avere questa parte come sezione: "graph as functors"  ha
 senso come remark separato e la restante parte della sezione si salda a quella precedente}

Set-theoretical definitions of graphs are straightforward, but a further level of abstraction allows us to highlight some interesting properties of these objects. We can indeed consider a graph as a functor from a category of ``primitive objects'' (e.g., the object representing edges and the object representing vertices) onto a category which provides a representation of them. This concept is perfectly represented by categories of presheaves, where an object is nothing more than a way to interpret objects of a category as sets. An example will make clear what we are talking about.

@@

    A right adjoint of $Q$ is the functor $T: \mathbf{Graph \rightarrow EqGrph}$ \todo{stesso problema di prima. Io riscriverei anche questa metà perché non si capisce la notazione e ad un certo punto occorre usare le proprietà dei quozienti in Set, cosa che non è evidente.} such that, for each graph $\mathcal{G}$ of $\mathbf{Graph}$, $T(\mathcal{G}) = (\mathcal{G}, \approx_{\mathcal{G}})$, where $\approx_{\mathcal{G}}$ is the total relation among vertices of $\mathcal{G}$. We have then the following situation:
    \[
        \begin{tikzcd}
            \eqgraph{G} = (\mathcal{G}, \sim_{\mathcal{G}}) \arrow[r, "{\eta_{\eqgraph{G}}}"] \arrow[dr, "f"swap] & (T \circ Q)(\eqgraph{G}) = (\mathcal{G}/_{\sim_{\mathcal{G}}}, \approx_{\mathcal{G}/_{\sim_{\mathcal{G}}}}) \arrow[d, dashed, "h"] \\
            & T(\mathcal{H}) = \eqgraph{H}
        \end{tikzcd}
    \]
    Since $\eqgraph{H}$ is in the image of $T$, it is a graph with all equivalent nodes. Then, the morphism $f$ is a graph morphism that in addition sends each equivalence class of vertices onto the unique class in the graph $\eqgraph{H}$. Then, $f$ factors uniquely as $h \circ \eta_\eqgraph{G}$, and $h$ is a graph morphism extended to a morphism in $\mathbf{EqGrph}$ by sending the unique equivalence class on $(T \circ G)(\eqgraph{G})$ onto the unique equivalence class in $\eqgraph{H}$, and $\eta: Id_{\mathbf{EqGrph}} \rightarrow (T \circ Q)$ is the unit of the adjunction.
\end{proof}

@@

\begin{proof}
    Consider the following pullback in $\mathbf{EqGrph}$, where $m$ is a regular monomorphism:
    \[
        \begin{tikzcd}
            G \arrow[r, "m"] \arrow[d, "n" swap] & G_1 \arrow[d, "i"] \\
            G_2 \arrow[r, "j" swap] & P
        \end{tikzcd}
    \]
    In particular, since $\mathbf{EqGrph}$ is a category of presheaves, colimits are computed pointwise, we have the following situation in the codomain category of the presheaves:
    \[
        \begin{tikzcd}
            G(Q) \arrow[r, "{m_Q}"] \arrow[d, "n_{Q}" swap] & G_1(Q) \arrow[d, "i_{Q}"] \\
            G_2(Q) \arrow[r, "j_{Q}" swap] & P(Q)
        \end{tikzcd}
    \]
    From \Cref{prop:monos_are_preserved_by_pullbacks_in_adh_cats}, the fact that $m$ is regular mono (i.e., $m=(m_E, m_V, m_Q)$ is mono on each component) implies that $j = (j_E, j_V, j_Q)$ is mono on each component, hence a regular mono in $\mathbf{EqGrph}$.

    %%% IDEA: far vedere che, poiché i colimiti in una categoria di prefasci sono calcolati punto punto, allora ho un pushout su ogni componente (E, V, Q). Dalla prop nella parte sull'adesività si può concludere che tutti e tre i pushout preservano mono e quindi mono è preservata su tutte e tre le componenti, quindi reg  mono.
\end{proof}
}\color{black}
\todo[color=green!40]{
COSE DA FARE
>Quoziente crea i limiti (questo dovrebbe derivare direttamente dalla sezione precedente ) 
>Quoziente crea certi pushout (questo va fatto ex novo)
}
% Regular monos in $\mathbf{EqGrph}$ are then natural transformations $\eta: \eqgraph{G \dot\rightarrow H}$ such that $\eta_Q$ is mono (\Cref{prop:reg_mono_in_EG_are_mono_in_graph}). Pullbacks preserve regular monomorphisms (\Cref{prop:monos_pres_by_pullback}). Regular monos are preserved also by pushouts (\Cref{lemma:reg_mono_pres_by_pushouts}). Again, since (co)limits are computed pointwise, and in $\Set$ monos are preserved by pushouts (since $\Set$ is adhesive), we can conclude what follows.

@@

\begin{lemma}
    The quotient functor creates pullbacks.
\end{lemma}

\begin{proof}
    We will show now that $Q$ reflects pullbacks.
    Consider the following diagram in $\mathbf{Graph}$
    \[
        \begin{tikzcd}
            & Q(\eqgraph{G}) \arrow[d, "Q(f)"] \\
            Q(\eqgraph{H}) \arrow[r, "Q(g)"swap] & Q(\eqgraph{K})
        \end{tikzcd}
    \]
    and let $(\mathcal P, p_{Q(f)}, p_{Q(g)})$ be the pullback. We want to show that, if  $(\eqgraph{P}, p_f, p_g)$ is the pullback of the arrows $f$ and $g$ in $\mathbf{EqGrph}$, then $Q(\eqgraph{P}) = \mathcal{P}$, $Q(p_f) = p_{Q(f)}$ and $Q(p_g) = p_{Q(g)}$. \color{green}{continuare}\color{black}
    Hence, from \Cref{cor:quot_preserves_co_lim}, we can conclude that $Q$ creates pullbacks.
\end{proof}

\begin{lemma}
    The quotient functor creates pullbacks along regular monomorphism.
\end{lemma}

@@

\todo[color=green!40]{
    COSE DA FARE:
Questa sezione ha più o meno gli stessi problemi della precedente. L'ordine da rispettare imho è il seguente:
>Definizione
>Calcolo dei limiti e certi colimiti (si fanno come in EqGrph)
> cor del calcolo dei limiti: caratterizzare mono regolari
>I crea limiti e i giusti pushout
> e-graph sono quasiadesivi
}
